In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-10 paper/main.tex
More updates in Methods.

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Beginning with the advective, net flow flield driven by 
CSF production, we impose a constant CSF inflow of $400\,$ml/day across the surface
of the lateral ventricles. Simultaneously allowing for efflux across the upper 
convexity yields a total CSF pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). Next, to examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Supplementary information S1.3 for a complete description). 
To estimate the cardiac contribution, we compute the steady state
CSF pressure and flow fields at peak systolic blood inflow.
Modelling the reduction in CSF space volume corresponding to peak systolic
conditions, we assume a total inflow rate of $6\,$ml/s~\cite{baledent2014imaging, causemann2022human}
in the SAS across its outer surface and $0.31\,$ml/s 
across the lateral ventricle surface\cite{vinje2019respiratory}. 
This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

@@

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--B, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Due to imaging limitiations, the network primarily covers the supratentorial regions, with the lowest network point located approximately $4\,$cm above the craniocervical junction.
Imposing the pressure field induced by CSF production at the periarterial network ends induces slow steady CSF flow of variable direction in these PVSs, with
an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as
the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep,
  broggini2024long,hauglund2025norepinephrine}, also induce net
directional flow in the PVS~\cite{kedarasetti2020arterial,
  kedarasetti2020functional, coenen2021lubrication,
  gjerde2023directional, nozaleda2024arterial} -- of magnitude and
direction depending on the amplitude, frequency, and length of the
waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a semi-analytic model
of the net flow induced by peristalsis in perivascular
networks~\cite{gjerde2023directional} (see Methods), we estimate that
the cardiac pulse wave alone, traveling at a frequency of 1 Hz with a
wavelength of 2.0 m and a 1\% wall displacement~\cite{jung2021novel},
will induce mainly antegrade PVS flow with an average net velocity of
0.92 \textmu m/s while reaching up to 7.31 \textmu m/s near larger,
ventral vessels such as the MCA (\Cref{fig:pvs_a}D,E). The same theory
predicts that strong ultraslow vasomotions, if traveling antegrade at
0.1Hz with a wavelength of 0.02 m and a 10\%
displacement~\cite{broggini2024long}, will induce both retrograde and
antegrade net PVS flow with an average velocity of 13.05 \textmu m/s
and maximum velocity 54.44 \textmu m/s (\Cref{fig:pvs_a}D,F).

@@

We represent the evolution and distribution of a solute in the CSF
spaces $\Omega_{\rm CSF}$, brain parenchyma $\Omega_{\rm PAR}$, and
periarterial and perivenous networks $\Lambda_a \cup \Lambda_v$ via a
mixed-dimensional transport model~\cite{masri2024modelling} over a timescale
of minutes to hours (\Cref{fig:results1}A--B, \Cref{fig:csf}A--C). The
current model accounts for solute exchange between the CSF spaces,
brain parenchyma, and the PVSs, diffusion within all compartments,
dispersion and convection in the CSF spaces, and convection within the
PVS. Specifically, for $t > 0$, we solve for a concentration field
$c(x, t)$ in the 3D intracranial compartments (for $x \in \Omega_{\rm
  CSF} \cup \Omega_{\rm PAR}$) and for a cross section-averaged
concentration field $\hat{c}(s, t)$ in the perivascular networks (for
$s \in \Lambda_a, \Lambda_v$) satisfying the following equations:
\begin{subequations}
\begin{alignat}{2}
  \partial_t c - \nabla \cdot D (1 + R) \nabla c  + \nabla \cdot (\bm u_{\rm CSF} c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm CSF},
  \label{eq:multi_transport_3d_CSF}
  \\ 
  \partial_t (\phi c) - \nabla \cdot D^{\ast} \nabla (\phi c)  + \nabla \cdot (\bm u_{\rm PAR} c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm{PAR}},
  \label{eq:multi_transport_3d_PAR}
  \\ 
  \partial_t (A  \hat c) - \partial_s(\hat D A \partial_s ( \hat c)) +\partial_s(A \hat u \hat c )  +  \xi P (\hat c - \overline{c})  &= 0 && \quad \quad \mathrm{in} \quad  \Lambda_a, \Lambda_v .
  \label{eq:multi_transport_1d}
 \end{alignat}%
\label{eq:multi_transport}%
\end{subequations}%
In~\eqref{eq:multi_transport_3d_CSF}, $D$ is the (free) diffusion
coefficient of the solute, $R$ is a dispersion coefficient, $\bm
u_{\rm CSF}$ is a CSF velocity, $\xi$ is a permeability allowing for
transfer/exchange across the interfaces $\Gamma_a, \Gamma_v$ between
the periarterial and perivenous networks and their surroundings, the
overlined $\overline{c}$ denotes the average of $c$ over
cross-sections of $\Gamma_a, \Gamma_v$, and $\delta_{\Gamma}$ is a
Dirac term concentrated on $\Gamma_a, \Gamma_v$. Additionally, in
\eqref{eq:multi_transport_3d_PAR}, $\phi$ is the brain extracellular
volume fraction\footnote{We set $\phi = 1.0$ in $\Omega_{\rm CSF}$ and
therefore omit $\phi$ in~\eqref{eq:multi_transport_3d_CSF}.},
$D^{\ast}$ is the effective diffusion coefficient in the parenchyma,
and $\bm u_{\rm PAR}$ is an interstitial fluid velocity. In this
study, we set $\bm u_{\rm PAR} = 0$. In~\eqref{eq:multi_transport_1d},
$A = \pi (R_2^2 - R_1^2)$ is the area and $P = 2 \pi R_2$ is the
perimeter of the PVS cross-sections, $\hat{D}$ is the effective
diffusion coefficient in the PVSs, and $\hat{u}$ is the fluid velocity
(in the axial direction) in the PVS. Further, we model the membrane
between the parenchyma and the CSF spaces as a semi-permeable membrane
with permeability coefficient $\beta_{\rm pia}$. The initial
concentrations in the three-dimensional domain $\Omega$ and networks
$\Lambda_a, \Lambda_v$ are all set to zero.

@@

scenarios. We assume that the porosity equals the extracellular space
volume fraction $\phi = \phi_{\rm PAR}$ within the parenchyma and
$\phi = 1$ elsewhere. The effective diffusion coefficient is set equal
to that of Gadubutrol in free water in the PVS networks: $\hat{D} =
D^{\rm Gad}$; weighted by a dispersion factor $R$ in the CSF spaces:
$D|_{\Omega_{\rm CSF}} = ( 1 + R) D^{\rm Gad}$; and modulated by the
porosity and tortuosity ($\lambda = 1.78$) in the parenchyma:
$D|_{\Omega_{\rm PAR}} = D^{\rm Gad}_{\rm PAR}$. We set $\xi$
depending on whether the PVS segment is fully embedded within the CSF
spaces ($\xi = \xi_{\rm CSF}$) or parenchyma ($\xi = \xi_{\rm EF}$),
or neither (see also Supplementary information S1). To account for the unresolved continuation of the perivascular network beyond the imaging resolution limit, we implemented a permeability gradient to avoid artificial tracer buildup. At the terminal 'leaf nodes' of $\Lambda_a$, permeability was increased by a factor of 100 to simulate tracer exchange with the surrounding CSF/tissue, before linearly decreasing to the baseline value at the adjacent network node.

@@

      Extracellular space volume fraction & $\phi_{\rm PAR}$ & 0.2 & -- & \cite{nicholson1981ion} \\

>>>>

For the CSF dynamics in our baseline model, we combine the net
contribution to CSF flow induced by CSF production with the integrated
dispersive effects of cardiac and respiratory pulsatile CSF flow
(\Cref{fig:csf}A--C). Beginning with the advective, net flow field driven by 
CSF production, we impose a constant CSF inflow of $400\,$ml/day across the surface
of the lateral ventricles. Simultaneously allowing for efflux across the upper 
convexity yields a total CSF pressure drop of $26\,$mPa (0.00020 mmHg) (\Cref{fig:csf}D) with a
maximum flow velocity of $1.85\,$mm/s (in the aqueduct) and a mean
velocity in the SAS of $2.6$\textmu m/s (\Cref{fig:csf}E). Next, to examine the
dispersive effects of pulsatile CSF flow, we employ the theory of
shear-augmented dispersion together with computational fluid dynamics
to determine a dispersion factor $R$ enhancing the effective solute
diffusivity (see Supplementary information S1.3 for a complete description). 
To estimate the cardiac contribution, we compute the steady state
CSF pressure and flow fields at peak systolic blood inflow.
Modelling the reduction in CSF space volume corresponding to peak systolic
conditions, we assume a total inflow rate of $6\,$ml/s~\cite{baledent2014imaging, causemann2022human}
in the SAS across its outer surface and $0.31\,$ml/s 
across the lateral ventricle surface\cite{vinje2019respiratory}. 
This scenario sets up a pressure
drop of $10\,$Pa ($0.075$ mmHg) between the lateral ventricles and the
spinal SAS and a maximum flow velocity of $19.8\,$cm/s
(\Cref{fig:csf}F, J). Assuming a cardiac frequency of $1\,$Hz, we
infer that this cardiac-induced pulsatile CSF flow increases the
effective diffusion by more than two orders of magnitude in the
aqueduct and near the cisterna magna, but has little effect ($R <
1$) in most of the SAS (\Cref{fig:csf}G, K). For the respiratory
contribution, we employ the same methodology, but with a total rate of
$1\,$ml/s \cite{gutierrez2022effect} in the SAS and $0.121\,$ml/s
\cite{liu2024using} in the lateral ventricles, yielding a respiratory
peak flow volume of $1.121\,$ ml/s at the craniocervical
junction. While the resulting flow velocities are only about one
fourth of their cardiac-induced counterparts (\Cref{fig:csf}H, L),
respiratory dispersive mixing reaches a factor of up to 320 due the
lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I, M).

@@

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--B, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Due to imaging limitations, the network primarily covers the supratentorial regions, with the lowest network point located approximately $4\,$cm above the craniocervical junction.
Imposing the pressure field induced by CSF production at the periarterial network ends induces slow steady CSF flow of variable direction in these PVSs, with
an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as
the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep,
  broggini2024long,hauglund2025norepinephrine}, also induce net
directional flow in the PVS~\cite{kedarasetti2020arterial,
  kedarasetti2020functional, coenen2021lubrication,
  gjerde2023directional, nozaleda2024arterial} -- of magnitude and
direction depending on the amplitude, frequency, and length of the
waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a semi-analytic model
of the net flow induced by peristalsis in perivascular
networks~\cite{gjerde2023directional} (see Methods), we estimate that
the cardiac pulse wave alone, traveling at a frequency of 1 Hz with a
wavelength of 2.0 m and a 1\% wall displacement~\cite{jung2021novel},
will induce mainly antegrade PVS flow with an average net velocity of
0.92 \textmu m/s while reaching up to 7.31 \textmu m/s near larger,
ventral vessels such as the MCA (\Cref{fig:pvs_a}D,E). The same theory
predicts that strong ultraslow vasomotions, if traveling antegrade at
0.1Hz with a wavelength of 0.02 m and a 10\%
displacement~\cite{broggini2024long}, will induce both retrograde and
antegrade net PVS flow with an average velocity of 13.05 \textmu m/s
and maximum velocity 54.44 \textmu m/s (\Cref{fig:pvs_a}D,F).

@@

We represent the evolution and distribution of a solute in the CSF
spaces $\Omega_{\rm CSF}$, brain parenchyma $\Omega_{\rm PAR}$, and
periarterial and perivenous networks $\Lambda_a \cup \Lambda_v$ via a
mixed-dimensional transport model~\cite{masri2024modelling} over a timescale
of minutes to hours (\Cref{fig:results1}A--B, \Cref{fig:csf}A--C). The
current model accounts for solute exchange between the CSF spaces,
brain parenchyma, and the PVSs, diffusion within all compartments,
dispersion and convection in the CSF spaces, and convection within the
PVS. Specifically, for $t > 0$, we solve for a concentration field
$c(x, t)$ in the 3D intracranial compartments (for $x \in \Omega_{\rm
  CSF} \cup \Omega_{\rm PAR}$) and for a cross section-averaged
concentration field $\hat{c}(s, t)$ in the perivascular networks (for
$s \in \Lambda_a, \Lambda_v$) satisfying the following equations:
\begin{subequations}
\begin{alignat}{2}
  \partial_t c - \nabla \cdot D (1 + R) \nabla c  + \nabla \cdot (\bm u_{\rm CSF} c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm CSF},
  \label{eq:multi_transport_3d_CSF}
  \\ 
  \partial_t (\phi c) - \nabla \cdot D^{\ast} \nabla (\phi c)  + \nabla \cdot (\bm u_{\rm PAR} c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm{PAR}},
  \label{eq:multi_transport_3d_PAR}
  \\ 
  \partial_t (A  \hat c) - \partial_s(\hat D A \partial_s ( \hat c)) +\partial_s(A \hat u \hat c )  +  \xi P (\hat c - \overline{c})  &= 0 && \quad \quad \mathrm{in} \quad  \Lambda_a, \Lambda_v .
  \label{eq:multi_transport_1d}
 \end{alignat}%
\label{eq:multi_transport}%
\end{subequations}%
In~\eqref{eq:multi_transport_3d_CSF}, $D$ is the (free) diffusion
coefficient of the solute, $R$ is the dispersion coefficient in the
CSF spaces, $\bm u_{\rm CSF}$ is the steady CSF velocity, $\xi$ is a
permeability allowing for transfer/exchange across the interfaces
$\Gamma_a, \Gamma_v$ between the periarterial and perivenous networks
and their surroundings, the overlined $\overline{c}$ denotes the
average of $c$ over cross-sections of $\Gamma_a, \Gamma_v$, and
$\delta_{\Gamma}$ is a Dirac term concentrated on $\Gamma_a,
\Gamma_v$. Additionally, in \eqref{eq:multi_transport_3d_PAR}, $\phi$
is the brain extracellular volume fraction\footnote{We set $\phi =
1.0$ in $\Omega_{\rm CSF}$ and therefore omit $\phi$
in~\eqref{eq:multi_transport_3d_CSF}.}, $D^{\ast}$ is the effective
diffusion coefficient in the parenchyma, and $\bm u_{\rm PAR}$ is an
interstitial fluid velocity. In this study, we set $\bm u_{\rm PAR} =
0$. In~\eqref{eq:multi_transport_1d}, $A = \pi (R_2^2 - R_1^2)$ is the
area and $P = 2 \pi R_2$ is the perimeter of the PVS cross-sections,
$\hat{D}$ is the effective diffusion coefficient in the PVSs, and
$\hat{u}$ is the fluid velocity (in the axial direction) in the PVS
defined for the periarterial and perivenous spaces separately
($\hat{u}|_{\Lambda_a} = \hat{u}_a$, $\hat{u}|_{\Lambda_v} =
\hat{u}_v$). Further, we model the membrane between the parenchyma and
the CSF spaces as a semi-permeable membrane with permeability
coefficient $\beta_{\rm pia}$, allowing for a jump in $c$ there. The
initial concentrations in $\Omega$ and networks $\Lambda_a, \Lambda_v$
are all set to zero.

@@

scenarios.
%
%We assume that the porosity equals the extracellular space
%volume fraction $\phi = \phi_{\rm PAR}$ within the parenchyma and
%$\phi = 1$ elsewhere.
%
The effective diffusion coefficient is set equal to that of Gadubutrol
in water in the PVS networks: $\hat{D} = D \equiv D^{\rm Gad}$; weighted by
the dispersion factor $R$ in the CSF spaces: $(1 + R) D$;
and modulated by the porosity and tortuosity ($\lambda = 1.78$) in the
parenchyma $D^{\ast} = D^{\rm Gad}_{\rm PAR}$. We set
$\xi$ depending on whether the PVS segment is fully embedded within
the CSF spaces ($\xi = \xi_{\rm CSF}$) or parenchyma ($\xi = \xi_{\rm
  EF}$), or neither (see also Supplementary information S1). To
account for the unresolved continuation of the perivascular network
beyond the imaging resolution limit, we implemented a permeability
gradient to avoid artificial tracer buildup. At the terminal leaf
nodes of $\Lambda_a$, permeability was increased by a factor of 100
to simulate tracer exchange with the surrounding CSF/tissue, and
linearly decreasing to the baseline value at the adjacent network
node.

@@

      %Extracellular space volume fraction & $\phi_{\rm PAR}$ & 0.2 & -- & \cite{nicholson1981ion} \\
      Extracellular space volume fraction & $\phi$ & 0.2 & -- & \cite{nicholson1981ion} \\
